Evolution of ORM Systems: Formal Foundations

### 2013-01-17 doc/description.tex
fixed name collisions in associaitons and properties + updated mapping + hidden find operations

\subsubsection{Mapping of Properties}
There are two kinds of properties - primitive property with cardinality equal to one is mapped to a column; second kind - collection with cardinality greater than one is mapped to table, thus there are two mapping cases. When the property is mandatory the $NOTNULL$ constraint is added to the column with property value. 
\begin{align*}
	\mathbf{\rho_p} : Property* \rightarrow Column* \\ \\
	\rho_{sp}(p) = \begin{cases}
		\inference{p = \varnothing}{\varnothing}\\
		\inference{|p| > 1}{\rho_{sp}(head(p)) \cup \rho_{sp}(tail(p))}\\\\
 		\inference{cardinality(p) = 1 \wedge !\;mandatory(p)
 			\\ \wedge \exists t : t = table(name(owningClass(p)), d) 
		 }{\begin{gathered}
	  		column(name(p), \rho_t(type(p)), \varnothing )
		\end{gathered}} \\ \\
  		\inference{cardinality(p) = 1 \wedge mandatory(p)
			 \\ \wedge \exists t : t = table(name(owningClass(p)), d)
			\\ \wedge \not instanciated(owningClass(p))
		}{\begin{gathered}
 	 		column(name(p), \rho_t(type(p)), NOTNULL ) \\ 
		\end{gathered}}
	\end{cases}\\
	p \lhd Property
\end{align*}
\begin{align*}
	\mathbf{\rho_p} : Property* \rightarrow Table* \\ \\
	\rho_{cp}(p) = \begin{cases}
		\inference{p = \varnothing}{\varnothing}\\
		\inference{|p| > 1}{\rho_{cp}(head(p)) \cup \rho_{cp}(tail(p))}\\\\
     	\inference{cardinality(p) > 1 \wedge !\;mandatory(p) \\ 
     		\wedge 	\exists t : t = table(name(owningClass(p)), d)
     		}{\begin{gathered}
     			table(name(p), primaryKey("Id"), \\ column("value", 			\rho_t(type(p)), \varnothing), \\ foreignKey(name(p), 			name(owningClass(p)), NOTNULL) \\
    	\end{gathered}}\\ \\
     	\inference{ cardinality(p) > 1 \wedge mandatory(p) \\ 
     		\wedge \exists t : t = table(name(owningClass(p)), d) 
     		}{\begin{gathered}  
     			table(name(p), primaryKey("Id"), \\ 
     			column("value", \rho_t(type(p)), NOTNULL), \\ 			foreignKey(name(p), name(owningClass(p)), NOTNULL)
    	 \end{gathered}}
 	\end{cases}\\
 	p \lhd Property
\end{align*}

\paragraph{Mapping of Associations} An association is mapped into database as a foreign key in table representing associated class or as a table in case the association cardinality is one to many or many to many. Thus there are two mapping cases as for properties.
\begin{align*}
	\mathbf{\rho_a} : Association* \rightarrow ForeignKey* \\
	\rho_a(a) = \begin{cases}
		\inference{a = \varnothing}{\varnothing}\\
		\inference{|a| > 1}{\rho_{a}(head(a)) \cup \rho_{a}(tail(a))}\\\\
		\inference{ startCardinality(a) = 0 \\ \wedge \exists t : t = table(name(owningClass(p)), d) \\ \wedge \exists u : u = table(reference(a)), d)
		}{
			foreignKey(name(a), name(t),  \varnothing) 
	 	}
  \\ \\
 	 \inference{ startCardinality(a) = 1 \\ \wedge \exists t : t = table(name(owningClass(p)), d) \\ \wedge \exists u : u = table(reference(a)), d)
 	 }{ 
		foreignKey(name(a), name(t),  NOTNULL)
	}
	 \end{cases}\\
	 a \lhd Association
\end{align*}
\begin{align*} 
	\mathbf{\rho_a} : Association* \rightarrow Table* \\
	\rho_a(a) = \begin{cases}
 		\inference{a = \varnothing}{\varnothing}\\
		\inference{|a| > 1}{\rho_{a}(head(a)) \cup \rho_{a}(tail(a))}\\\\
		\inference{  startCardinality > 1 \\ \wedge \rho(owningClass(p)) \in tables(d) \wedge \rho(reference(a)) \in tables(d)
  		}{\begin{gathered}  
		 table(name(a), pk("Id"), \varnothing, \\ foreignKey(owningClass(a),owningClass(a), NOTNULL) \\ foreignkey(reference(a), reference(a), NOTNULL)) 
  		\end{gathered}}  
 	\end{cases}\\
	 a \lhd Association
\end{align*}

@@

\paragraph{Get Property or Association} Functions $findProperty$ and $findAssociation$ returns a property (association) if it exists in given context - application and class of provided name.
\begin{align*}
	\mathbf{findProperty} : Label \times Label \times Application \rightarrow Property \\
	findProperty(l_c, l_p, a) = \inference{\exists c \in classes(a): name(c) = l_c \\ \wedge \exists p \in properties(c) : name(p) = l_p}{p} \\
	l_c, l_p \in Label, a \lhd Application
\end{align*}
\begin{align*}
	\mathbf{findAssociation} : Label \times Label \times Application \rightarrow Association \\
	findAssociation(l_c, l_a, a) = \inference{\exists c \in classes(a): name(c) = l_c \\ \wedge \exists o \in associations(c) : name(o) = l_a}{o} \\
	l_c, l_a \in Label, a \lhd Application
\end{align*}

\paragraph{Add Property} The name collision only has to be verified, before adding a property into an application structure:
\begin{align*}
 	\mathbf{addProp} : Label \times Property \times Application \rightarrow Application \\
	addProp(l, p, a) = \inference{\exists c \in classes(a) : name(c) = l \\ \wedge \forall q \in  properties(p) : name(q) \neq name(p)}{properties(c) \cup p } \\
	l \in Label, p \lhd Property, a \lhd Applications
\end{align*}

\paragraph{Add Association} Adding an association into an application structure has to verify if there is no name collision and if there is the associated class present in the application model.
\begin{align*}
	\mathbf{addAssoc} : Label \times Association \times Application \rightarrow Application \\
	addAssoc(l, s, a) = \inference{\forall f \in associations(a) : name(f) \neq name(s) \\ \wedge \exists c \in classes(a) : name(c) = l \\ \wedge \exists k \in classes(a) : name(k) = reference(s) }{associations(c) \cup s}\\
	l \in Labels, s \in Associations, a \lhd Applications
\end{align*}

@@

\paragraph{Find Table} The function $findTable$ finds and returns a table in the database according to the provided label.
\begin{align*}
	\mathbf{findTable} : Label \times Database \rightarrow Table  \\
	findTable(l, d) = \inference{\exists t \in tables(d) \wedge name(t) = l}{t}  \\ \\
	l \in Label, d \lhd Database
\end{align*}

@@

\subsection{Software Decomposition}
The removal of class, property or association is limited only by the constraints given in application or database transformations. The main constraint is there has to be no data, which can be lost during removal. 
\paragraph{Remove class}
\begin{align*}
	\mathbf{RemoveClass}: Label \times Software \rightarrow State  \\
	\Psi RemoveClass(l, s) = applications(s) \circ remClass(l) \\
	\Phi RemoveClass(l, s) = database(s) \bullet dropTable(l) \\
	l \in Label, s \in Software 
\end{align*}
\paragraph{Remove property}
\begin{align*}
	\mathbf{RemoveProperty} : Label \times Label \times State \rightarrow State \\
	\Psi RemoveProperty(l, l_p, s) =  applications(s) \circ  remProp(l, l_p) \\
	\Phi RemoveProperty(l, l_p, s) = \begin{cases}
		 \inference{cardinality(findProperty(l_p, application(s)) \leq 1}{database(s) \bullet dropColumn(l, l_p)}  \\\\
		 \inference{cardinality(findProperty(l_p, application(s)) >1}{database(s) \bullet dropTable(l)}
 	\end{cases}\\
 	l, l_p \in Label, s \lhd Software
\end{align*}
\paragraph{Remove association} 
\begin{align*}
    \mathbf{RemoveAssociation} : Label \times Labels \times State \rightarrow State \\ 
    \Psi RemoveAssociation(l, l_a, s) = application(s) \circ remAss(l, l_a) \\
    \Phi RemoveAssociation(l, l_a, s) = \begin{cases}
     	\inference{a = findAssociation(l, l_a, application(s)) \\ \wedge startCardinality(a) \leq 1 \\}{database(s) \bullet dropForeignKey(l_a, reference(a))} \\\\
     	\inference{a = findAssociation(l, l_a, application(s)) \\ \wedge startCardinality(a) > 1}{database(s) \bullet dropTable(l_a)} 
     \end{cases}\\
    l_c, l_a \in Labels, s \lhd Software
\end{align*}

@@

\paragraph{Rename Property}
\begin{align*}
	\mathbf{RenameProperty} : Label \times Label \times Label \times State \rightarrow State \\ \\
\Psi(RenameProperty(l_c, l_o, l_n, s)) = application(s) \circ renProp(l_c, l_o, l_n) \\
\Phi(RenameProperty(l_c, l_o, l_n, s)) = \begin{cases}
 	\inference{cardinality(findProperty(l_c, l_o, application(s))) \leq 1}{database(s) \bullet alterColumn(l_c, l_o, l_n))} \\\\
 	\inference{cardinality(findProperty(l_c, l_o, application(s))) > 1}{database(s) \bullet alterTable(l_o, l_n, database(s))}
 \end{cases}\\
l_c, l_o, l_n \in Label,  s \lhd Software
\end{align*}

\paragraph{Rename Association}
\begin{align*}
\mathbf{RenameAssociation} : Label \times Label \times Label \times State \rightarrow State \\
\Psi(RenameAssocaiation(l_c, l_o, l_n), s) = application(s) \circ renAssoc(l_c, l_o, l_n)\\
\Phi(RenameAssocaiation(l_c, l_o, l_n), s) = \begin{cases}
	\inference{\exists a  \in findAssociation(application(s)) : \\
  name(a) = l_c \\ \wedge startCardinality(a) > 1 }{database(s) \bullet alterTable(l_c, l_o, l_n)}
 \\ \\
  \inference{\exists a  \in findAssociation(application(s)) : \\
  name(a) = l_c \\ \wedge startCardinality(a) \leq 1 }{\begin{gathered}
database(s) \\ \bullet alterForeignKey(endClass(a) ,l_c, l_o, l_n)
\end{gathered}}
 \end{cases}
\\\\
%	alterAssoc(l_c, l_o, l_n, d) = \begin{cases}
%?? \inference{\exists t \in tables(a) : name(t) = l_c \\ \wedge \exists f \in fks(t) : name(c) = l_o \\ \wedge \forall e \in fks(t) : name(e) \neq l_n}{\begin{gathered}
%f.name = l_n 
%\end{gathered}}
%\\ \\
%\inference{\exists t \in tables(a) : name(t) = l_o \\ \wedge \forall e \in tables(t) : name(e) \neq l_n}{\begin{gathered}
%t.name = l_n 
%\end{gathered}}
% \end{cases}
% \\
l_c, l_o, l_n \in Label, a \lhd Applications, s \in Sates, d \lhd Database
\end{align*}

>>>>

\subsubsection{Mapping of Properties}
There are two kinds of properties - primitive property with cardinality equal to one is mapped to a column; second kind - collection with cardinality greater than one is mapped to table, thus there are two mapping cases. When the property is mandatory the $NOTNULL$ constraint is added to the column with property value.

There can occur name collision during mapping in case there are more classes with property representing collection. To avoid that the mapping changes the name of the table representing a collection.
\begin{align*}
	\mathbf{\rho_p} : Property* \rightarrow Column* \\ \\
	\rho_{sp}(p) = \begin{cases}
		\inference{p = \varnothing}{\varnothing}\\
		\inference{|p| > 1}{\rho_{sp}(head(p)) \cup \rho_{sp}(tail(p))}\\\\
 		\inference{cardinality(p) = 1 \wedge !\;mandatory(p)
 			\\ \wedge \exists t : t = table(name(owningClass(p)), d) 
		 }{\begin{gathered}
	  		column(name(p), \rho_t(type(p)), \varnothing )
		\end{gathered}} \\ \\
  		\inference{cardinality(p) = 1 \wedge mandatory(p)
			 \\ \wedge \exists t : t = table(name(owningClass(p)), d)
			\\ \wedge \not instanciated(owningClass(p))
		}{\begin{gathered}
 	 		column(name(p), \rho_t(type(p)), NOTNULL ) \\ 
		\end{gathered}}
	\end{cases}\\
	p \lhd Property
\end{align*}
\begin{align*}
	\mathbf{\rho_p} : Property* \rightarrow Table* \\ \\
	\rho_{cp}(p) = \begin{cases}
		\inference{p = \varnothing}{\varnothing}\\
		\inference{|p| > 1}{\rho_{cp}(head(p)) \cup \rho_{cp}(tail(p))}\\\\
     	\inference{cardinality(p) > 1 \wedge !\;mandatory(p) \\ 
     		\wedge 	\exists t : t = table(name(owningClass(p)), d)
     		}{\begin{gathered}
     			table(name(t)+name(p), primaryKey("Id"), \\ column("value", \rho_t(type(p)), \varnothing), \\ foreignKey(name(p), 			name(owningClass(p)), NOTNULL) \\
    	\end{gathered}}\\ \\
     	\inference{ cardinality(p) > 1 \wedge mandatory(p) \\ 
     		\wedge \exists t : t = table(name(owningClass(p)), d) 
     		}{\begin{gathered}  
     			table(name(t)+name(p), primaryKey("Id"), \\ 
     			column("value", \rho_t(type(p)), NOTNULL), \\ 			foreignKey(name(p), name(owningClass(p)), NOTNULL)
    	 \end{gathered}}
 	\end{cases}\\
 	p \lhd Property
\end{align*}

\paragraph{Mapping of Associations} An association is mapped into database as a foreign key in table representing associated class or as a table in case the association cardinality is one to many or many to many. Thus there are two mapping cases as for properties and the same solution is used to avoid name collisions.
\begin{align*}
	\mathbf{\rho_a} : Association* \rightarrow ForeignKey* \\
	\rho_a(a) = \begin{cases}
		\inference{a = \varnothing}{\varnothing}\\
		\inference{|a| > 1}{\rho_{a}(head(a)) \cup \rho_{a}(tail(a))}\\\\
		\inference{ startCardinality(a) = 0 \\ \wedge \exists t : t = table(name(owningClass(p)), d) \\ \wedge \exists u : u = table(reference(a)), d)
		}{
			foreignKey(name(a), name(t),  \varnothing) 
	 	}
  \\ \\
 	 \inference{ startCardinality(a) = 1 \\ \wedge \exists t : t = table(name(owningClass(p)), d) \\ \wedge \exists u : u = table(reference(a)), d)
 	 }{ 
		foreignKey(name(a), name(t),  NOTNULL)
	}
	 \end{cases}\\
	 a \lhd Association
\end{align*}
\begin{align*} 
	\mathbf{\rho_a} : Association* \rightarrow Table* \\
	\rho_a(a) = \begin{cases}
 		\inference{a = \varnothing}{\varnothing}\\
		\inference{|a| > 1}{\rho_{a}(head(a)) \cup \rho_{a}(tail(a))}\\\\
		\inference{  startCardinality > 1 \\ \wedge \exists t \in tables(d) : name(owningClass(p)) = name(t) \\ \wedge \exists e \in tables(d) : name(reference(a)) = name(e)
  		}{\begin{gathered}  
		 table(name(t)+name(a), pk("Id"), \varnothing, \\ foreignKey(owningClass(a),owningClass(a), NOTNULL) \\ foreignkey(reference(a), reference(a), NOTNULL)) 
  		\end{gathered}}  
 	\end{cases}\\
	 a \lhd Association
\end{align*}

@@

%\paragraph{Get Property or Association} Functions $findProperty$ and $findAssociation$ returns a property (association) if it exists in given context - application and class of provided name.
%\begin{align*}
%	\mathbf{findProperty} : Label \times Label \times Application \rightarrow Property \\
%	findProperty(l_c, l_p, a) = \inference{\exists c \in classes(a): name(c) = l_c \\ \wedge \exists p \in properties(c) : name(p) = l_p}{p} \\
%	l_c, l_p \in Label, a \lhd Application
%\end{align*}
%\begin{align*}
%	\mathbf{findAssociation} : Label \times Label \times Application \rightarrow Association \\
%	findAssociation(l_c, l_a, a) = \inference{\exists c \in classes(a): name(c) = l_c \\ \wedge \exists o \in associations(c) : name(o) = l_a}{o} \\
%	l_c, l_a \in Label, a \lhd Application
%\end{align*}

\paragraph{Add Property} The name collision only has to be verified, before adding a property into an application structure:
\begin{align*}
 	\mathbf{addProp} : Label \times Property \times Application \rightarrow Application \\
	addProp(l, p, a) = \inference{\exists c \in classes(a) : name(c) = l \\ \wedge \forall q \in  properties(c) : name(q) \neq name(p) \\ \wedge \forall e \in associations(c) : name(e) \neq name(p)}{properties(c) \cup p } \\
	l \in Label, p \lhd Property, a \lhd Applications
\end{align*}

\paragraph{Add Association} Adding an association into an application structure has to verify if there is no name collision and if there is the associated class present in the application model.
\begin{align*}
	\mathbf{addAssoc} : Label \times Association \times Application \rightarrow Application \\
	addAssoc(l, s, a) = \inference{
	\exists c \in classes(a) : name(c) = l \\
	\wedge \forall f \in associations(c) : name(f) \neq name(s) \\ 	\wedge \forall p \in properties(c) : name(p) \neq name(s) \\ 
	\wedge \exists k \in classes(a) : name(k) = reference(s)}{associations(c) \cup s}\\
	l \in Labels, s \in Associations, a \lhd Applications
\end{align*}

@@

%\paragraph{Find Table} The function $findTable$ finds and returns a table in the database according to the provided label.
%\begin{align*}
%	\mathbf{findTable} : Label \times Database \rightarrow Table  \\
%	findTable(l, d) = \inference{\exists t \in tables(d) \wedge name(t) = l}{t}  \\ \\
%	l \in Label, d \lhd Database
%\end{align*}

@@

\subsection{Software Decomposition}
The removal of class, property or association is limited only by the constraints given in application or database transformations. The main constraint is there has to be no data, which can be lost during removal. 
\paragraph{Remove class}
\begin{align*}
	\mathbf{RemoveClass}: Label \times Software \rightarrow State  \\
	\Psi RemoveClass(l, s) = applications(s) \circ remClass(l) \\
	\Phi RemoveClass(l, s) = database(s) \bullet dropTable(l) \\
	l \in Label, s \in Software 
\end{align*}

\paragraph{Remove property}
\begin{align*}
	\mathbf{RemoveProperty} : Label \times Label \times State \rightarrow State \\
	\Psi RemoveProperty(l, l_p, s) =  applications(s) \circ  remProp(l, l_p) \\
	\Phi RemoveProperty(l, l_p, s) = \begin{cases}
		 \inference{\exists c \in classes(application(s)) : name(c) = l\\ 
		\wedge \exists p \in properties(c) : name(p) = l \\ \wedge cardinality(p) \leq 1}{database(s) \bullet dropColumn(l, l_p)}  \\\\
		 \inference{\exists c \in classes(application(s)) : name(c) = l\\ 
		\wedge \exists p \in properties(c)  : name(p) = l \\ \wedge cardinality(p)>1}{database(s) \bullet dropTable(name(owningClass(p)+l)}
 	\end{cases}\\
 	l, l_p \in Label, s \lhd Software
\end{align*}

\paragraph{Remove association} 
\begin{align*}
    \mathbf{RemoveAssociation} : Label \times Labels \times State \rightarrow State \\ 
    \Psi RemoveAssociation(l, l_a, s) = application(s) \circ remAss(l, l_a) \\
    \Phi RemoveAssociation(l, l_a, s) = \begin{cases}
     	\inference{\exists c \in classes(application(s)) : name(c) = l \\
     	\wedge \exists a \in associaions(c) : name(a) = l_a  \\
     	\wedge startCardinality(a) \leq 1 \\}{database(s) \bullet dropForeignKey(l_a, reference(a))} \\\\
     	\inference{\exists c \in classes(application(s)) : name(c) = l \\
     	\wedge \exists a \in associaions(c) : name(a) = l_a  \\
     	\wedge startCardinality(a) > 1}{database(s) \bullet dropTable(name(owningClass(a))+l_a)} 
     \end{cases}\\
    l_c, l_a \in Labels, s \lhd Software
\end{align*}

@@

\paragraph{Rename Property}
\begin{align*}
	\mathbf{RenameProperty} : Label \times Label \times Label \times State \rightarrow State \\ \\
\Psi(RenameProperty(l_c, l_o, l_n, s)) = application(s) \circ renProp(l_c, l_o, l_n) \\
\Phi(RenameProperty(l_c, l_o, l_n, s)) = \begin{cases}
 	\inference{\exists c \in classes(application(s)) : name(c) = l\\ 
		\wedge \exists p \in properties(c) : name(p) = l_o \\
 	cardinality(p) \leq 1}{database(s) \bullet alterColumn(l_c, l_o, l_n))} \\\\
 	\inference{\exists c \in classes(application(s)) : name(c) = l\\ 
		\wedge \exists p \in properties(c) : name(p) = l_o \\
 	cardinality(p) > 1}{\begin{gathered}
		database(s)\\ \bullet alterTable(name(owningClass(p)+l_o, l_n, database(s))
\end{gathered}}
 \end{cases}\\
l_c, l_o, l_n \in Label,  s \lhd Software
\end{align*}

\paragraph{Rename Association}
\begin{align*}
\mathbf{RenameAssociation} : Label \times Label \times Label \times State \rightarrow State \\
\Psi(RenameAssocaiation(l_c, l_o, l_n), s) = application(s) \circ renAssoc(l_c, l_o, l_n)\\
\Phi(RenameAssocaiation(l_c, l_o, l_n), s) = \begin{cases}
	\inference{\exists c \in classes(application(s)) : name(c) = l\\ 
		\wedge \exists a \in associations(c) : name(a) = l_o 
	\\ \wedge startCardinality(a) > 1 }{\begin{gathered}
		database(s) \\ \bullet alterTable(name(owningClass(a)) + l_o, l_n)
\end{gathered}}\\ \\
  \inference{\exists c \in classes(application(s)) : name(c) = l\\ 
		\wedge \exists a \in associations(c) : name(a) = l_o 
	\\ \wedge startCardinality(a) \leq 1 }{\begin{gathered}
database(s) \\ \bullet alterForeignKey(endClass(a) ,l_c, l_o, l_n)
\end{gathered}}
 \end{cases} \\
	l_c, l_o, l_n \in Label, a \lhd Applications, s \in Sates, d \lhd Database
\end{align*}
